---
title: "Advanced-R"
author: "Yasser"
date: "`r Sys.Date()`"
output: html_document
---

# Advanced R

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
getwd()
```

## Names and values

**The distinction between names and values, and when R will copy an object.**

Question 1

Given the following data frame, how do I create a new column called “3” that contains the sum of 1 and 2? You may only use $, not [[. What makes 1, 2, and 3 challenging as variable names?

```{r}
df <- data.frame(runif(3), runif(3))
names(df) <- c(1,2)
df
df$`3` <- df$`1` + df$`2` # needs the backticks to work 

df
```


```{r}
# We’ll use the lobstr package to dig into the internal representation of R objects.
#install.packages("lobstr")
library(lobstr)
```

### Binding basics

Consider the code:

```{r}
x <- c(1, 2, 3)
```

Accurate to say:

It’s creating an object, a vector of values, c(1, 2, 3).
And it’s binding that object to a name, x.

The object, or value, doesn’t have a name; it’s actually the name that has a value.

The name binds or references a value, which here is the vector.

Think of a name as a reference to a value. For example the code below doesn't create another copy of c(1, 2, 3), you actually get another binding to the existing object:

```{r}
y <- x
```


You can access an object’s identifier with lobstr::obj_addr(). Doing so allows you to see that both x and y point to the same identifier:

```{r}
obj_addr(x)

obj_addr(y)
# they have the same identifier - which indicates where in memory the value is stored.
```
These identifiers are long, and change every time you restart R.

**It can take some time to get your head around the distinction between names and values, but understanding this is really helpful in functional programming where functions can have different names in different contexts.**

### Non-syntactic names

R has rules around what constitutes a valid name. A **syntactic** name must consist of letters, digits, `.` and `_` but can't begin with `_` or a digit. In addition you can't use any of the reserved words like `TRUE`, `NULL`, `if` and `function`. There is a complete list (`?Reserved`). A name that doesn't follow these rules is non-syntactic and will cause an error:

```{r}
_abc <- 1
```

```{r}
if <- 10
```

It is possible to use any name by using backticks:

```{r}
`_abc` <- 1
`_abc`
`if` <- 10
`if`
```
Such instances may occur when loading data created outside of R.

**You can also create non-syntactic bindings using single or double quotes (e.g. "_abc" <- 1) instead of backticks, but you shouldn’t, because you’ll have to use a different syntax to retrieve the values. The ability to use strings on the left hand side of the assignment arrow is an historical artefact, used before R supported backticks.**

#### Exercises

1. Explain the relationship between a, b, c and d in the following code:

```{r}
a <- 1:10
b <- a
c <- b
d <- 1:10
```


```{r}
obj_addr(a)
obj_addr(b)
obj_addr(c)
obj_addr(d)

# a, b and c are all the same - they are referencing the same object, as confirmed by the objects address
# d has created a new object, confirmed by the different object address
```

2. The following code accesses the mean function in multiple ways. Do they all point to the same underlying function object? Verify this with lobstr::obj_addr().

```{r}
mean
base::mean
get("mean")
evalq(mean)
match.fun("mean")

# all the above point to the same underlying function, as confirmed by the the address associated with each (bytecode below)
# They are also all in the same environment
```

3. base R data import functions, like read.csv(), will automatically convert non-syntactic names to syntactic ones. Why might this be problematic? What option allows you to suppress this behaviour?

```{r}
?read.csv
# the check.names argument:
# logical. If TRUE then the names of the variables in the data frame are checked to ensure that they are syntactically valid variable names. If necessary they are adjusted (by make.names) so that they are, and also to ensure that there are no duplicates.
# this could be an problem if it resulsts in information about the variable being lost.
```

4. What rules does make.names() use to convert non-syntactic names into syntactic ones?

```{r}
?make.names
# A syntactically valid name consists of letters, numbers and the dot or underline characters and starts with a letter or the dot not followed by a number. Names such as ".2way" are not valid, and neither are the reserved words.
# 
# The definition of a letter depends on the current locale, but only ASCII digits are considered to be digits.
# 
# The character "X" is prepended if necessary. All invalid characters are translated to ".". A missing value is translated to "NA". Names which match R keywords have a dot appended to them. Duplicated values are altered by make.unique.
```

5. Why is .123e1 not a syntactic name? -- as above

### Copy on modify

Consider the following code. It binds x and y to the same underlying value, then modifies y

```{r}
x <- c(1, 2, 3)
y <- x

y[[3]] <- 4
print(x)

print(y)
```

Modifying y clearly didn’t modify x. So what happened to the shared binding? While the value associated with y changed, the original object did not. Instead, R created a new object, 0xcd2, a copy of 0x74b with one value changed, then rebound y to that object.

This is referred to as **copy-on-modify**. **Understanding this concept significantly improves understanding the performance of R code.** With copy on modify you get different results in R Studio, as the environment pane makes a reference to each object to display information about it. For experimentation here it was therefore suggested to use R Markdown or the terminal.

#### tracemem()

You can see when an object gets copied with the help of base::tracemem(). Once you call that function with an object, you’ll get the object’s current address:

```{r}
x <- c(1, 2, 3)
cat(tracemem(x), "\n")

```
Now, whenever the object is copied, `tracemem()` will print a message telling you which object was copied, its new address, and the sequence of calls that led to the copy:

```{r}
y <- x
y[[3]] <- 4L
```
If we modify y again, it doesn't get copied. This is as the new object now only has a single name bound to it, so R applies **modify-in-place optimisation.**

```{r}
y[[3]] <- 5L

# untracemem() turns tracing off:
untracemem(x)
```


```{r}
original <- c(10, 20, 30, 40, 50)
cat(tracemem(original), "\n")

copy <- original
copy[[5]] <- 100


```
```{r}
copy
copy[[5]] <- 200
untracemem(copy)
```


#### Function calls

The same rules for copying also apply to function calls. Take this code:

```{r}
f <- function(a) {
  a
}

x <- c(1, 2, 3)
cat(tracemem(x), "\n")

z <- f(x)
# there is no copy here

untracemem(x)
```

While f() is running, the a inside the function points to the same value as the x does outside the function.

Once f() completes, x and z point to the same object. The object (the value) never gets copied because it never gets modified. If f() did modify x, R would create a new copy, and then z would bind that object.

#### Lists

It isn't just variables that point to values, elements of lists do too. 

```{r}
l1 <- list(1, 2, 3)
```

The lists stores references to each value.
When we copy a list as below:

```{r}
l2 <- l1
```

It references the same values as the list it copied.
When we modify a list:

```{r}
l2[[3]] <- 4
l2
```
The first two values above, reference the same two values as l1. However, the third value above (4) now references another value, not referenced by the first list.

Like vectors, lists use copy-on-modify behaviour; the original list is left unchanged, and R creates a modified copy. This, however, is a **shallow copy**: the list object and its bindings are copied, but the values pointed to by the bindings are not. The opposite of a shallow copy is a deep copy where the contents of every reference are copied. Prior to R 3.1.0, copies were always deep copies.

We can use `lobstr::ref` to see values that are shared across lists. `ref()` prints the memory address of each object as well as a local id. This helps identify shared components.

```{r}
ref(l1, l2)
```
Above we can see the two values that are shared across both lists.

#### Data frames

Data frames are lists of vectors, so copy-on-modify has important consequences when you modify a data frame. For example:

```{r}
d1 <- data.frame(x = c(1, 5, 6), y = c(2, 4, 3))
d1
```

If we modify a column, only that column needs to modified, the others will point to their original references.

```{r}
d2 <- d1
d2[, 2] <- d2[, 2] * 2
d2
```

Above, column X still points to the same column from d1. Although column y now points to the other modified vector of values.
However, if we modify a row, every column is changed, which means every column must be copied.

```{r}
d3 <- d1
d3[1, ] <- d3[1, ] * 3
d3
```

SO now the above data frame references two different vectors compared to d1, whereas as d2 shared a column with d1. This is confirmed below...

```{r}
ref(d1, d2, d3)
```

#### Character vectors

The final place that R uses references is with character vectors.

```{r}
x <- c("a", "a", "abc", "d")
```

R actually uses a global string pool where each element of a character vector is a pointer to a unique string in the pool.

We can use the `ref()` function to and set the `character` argument to `TRUE` to see these refernces.

```{r}
ref(x, character = TRUE)
```
We can see the the first two characters in the vector reference the same string in the global string pool. This is important when considereing the amount of memory a character vector uses.

#### Exercises

tracemem(1:10) is not useful as no object is being referenced, or stored in memory, so we then can't track it.

```{r}
tracemem(1:10)
```
```{r}
x <- c(1L, 2L, 3L)
tracemem(x)

x[[3]] <- 4
```

```{r}
a <- 1:10
b <- list(a, a)
c <- list(b, a, 1:10)
a
b
c
ref(b, c)
```

```{r}
x <- list(1:10)
x[[2]] <- x
x
ref(x)
```

### Obeject size

Can check the size of an object with `lobstr::obj_size()`

```{r}
obj_size(letters)
obj_size(ggplot2::diamonds)
```
Since the elements of lists are references to values, the size of a list are smaller than you expect:

```{r}
x <- runif(1e6)
obj_size(x)

y <- list(x, x, x)
obj_size(y)
```
y is only 80 bytes^14 bigger than x. That’s the size of an empty list with three elements:

```{r}
obj_size(list(NULL, NULL, NULL))
```
This is because y just references x three times, it doesn't copy the values.

Similarly, because R uses a global string pool character vectors take up less memory than you might expect: repeating a string 100 times does not make it take up 100 times as much memory.

```{r}
banana <- "bananas bananas bananas"
obj_size(banana)

obj_size(rep(banana, 100))
```
References also add complexity when thinking about the size of individual objects. obj_size(x) + obj_size(y) will only equal obj_size(x, y) if there are no shared values. Below, the combined size of x and y is the same as the size of y:

```{r}
obj_size(x, y)
```

Finally, R 3.5.0 and later versions have a surprising feature: ALTREP, **alternative representation**. This allows R to represent certain types of vectors very compactly. The place you are most likely to see this is with : because instead of storing every single number in the sequence, R just stores the first and last number. This means that every sequence, no matter how large, is the same size:

```{r}
obj_size(1:3)
#> 680 B
obj_size(1:1e3)
#> 680 B
obj_size(1:1e6)
#> 680 B
obj_size(1:1e9)
#> 680 B
```

#### Exercises

```{r}
y <- rep(list(runif(1e4)), 100)

object.size(y)
#> 8005648 bytes
obj_size(y)
#> 80,896 B
```
```{r}
?object.size # does not detect if elements of a list are shared and therefore gives a larger estimate of object size
```

```{r}
funs <- list(mean, sd, var)
obj_size(funs)
#> 17,608 B
```
```{r}
a <- runif(1e6)
obj_size(a) # 8,000,048 B

b <- list(a, a)
obj_size(b) # 8,000,112 B only slightly larger than original as pointing to a
obj_size(a, b) # same size as B

#b

b[[1]][[1]] <- 10

obj_size(b) # 16,000,160 B
obj_size(a, b) # 16,000,160 B

b[[2]][[1]] <- 10
obj_size(b) # 16,000,160 B
obj_size(a, b) # 24,000,208 B
```
### Modify in place

As seen, modifying an object usually creates a copy. There are two exceptions:

 - Objects with a single binding get a special performance optimisation.
 - Environments, a special type of object, are always modified in place.
 
#### Objects with a single binding

**If an object has single name bound to it, R will modify it in place.

```{r}
v <- c(1, 2, 3)

v[[3]] <- 4

```

Above, the same object will be modified, no copy will be made.

**Note: that above, the function `tracemem()` will reveal a copy is being made here.** As below:

```{r}
v <- c(1, 2, 3)
tracemem(v)
v[[3]] <- 4
tracemem(v)
```

```{r}
untracemem(v)
```

This is apparently due to this code being run in R Studio. As described here: https://stackoverflow.com/questions/15559387/operator-in-rstudio-and-r

"Rstudio's object browser modifies objects it examines in a way that forces copying upon modification."

**End note**

Two complications make predicting exactly when R applies this optimisation challenging:

 - With regards to bindings, R currently counts 0, 1 or many. Therefore if an object have two bindings, and one is removed, the reference count does not go back to 1. Therefore, R will make copies when it sometimes doesn't need to.
 - Whenever you call the vast majority of functions, it makes a reference to the object. The only exception are specially written “primitive” C functions. These can only be written by R-core and occur mostly in the base package.
 
Thus is can be difficult to predict whether a copy will be made, so it is best to use `tracemem()`

We can explore these issues with a for loop - the slowness of which in R is caused by every iteration of the loop creating a copy. The following subtracts the median from each column of a data frame:

```{r}
x <- data.frame(matrix(runif(5 * 1e4), ncol = 5))
medians <- vapply(x, median, numeric(1))

for (i in seq_along(medians)) {
  x[[i]] <- x[[i]] - medians[[i]]
}

```

This loop is slow because each iteration of the loop copies the data frame.

```{r}
cat(tracemem(x), "\n")

for (i in 1:5) {
  x[[i]] <- x[[i]] - medians[[i]]
}
```
```{r}
untracemem(x)
```

The number of copies can be reduced by using a list instead of a data frame. Modifying a list uses internal C code.

```{r}
y <- as.list(x)
cat(tracemem(y), "\n")
  
for (i in 1:5) {
  y[[i]] <- y[[i]] - medians[[i]]
}

#tracemem[0x000001b631501bc8 -> 0x000001b631f6f428]: 
```
*Preventing copies from being made can be tricky*

#### Environments

**Environments are always modified in place.** This sometimes referred to as **reference semantics** because when you modify an environment, all existing bindings to that environment will continue to have the same reference.

We bind the following environment to e1 and e2.

```{r}
e1 <- rlang::env(a = 1, b = 2, c = 3)
e2 <- e1
e1
e2
```

If we change a binding, the environment is modified in place:

```{r}
e1$c <- 4
# above we changed the e1 environment binding of c. Due to modify in place, e2 has has the same change:
e2$c
#4
```
Due to this property, environments can contain themselves:

```{r}
e <- rlang::env()
e$self <- e

ref(e)

# [1:0x1b632df1420] <env> 
#\-self = [1:0x1b632df1420] 
```

### Unbinding and the garbage collector

Example code:

```{r}
x <- 1:3

x <- 2:4

rm(x)

x
```
Following the creation of two objects, these objects are no longer bound to a name at the end of the chunk as x is `rm()`

The objects are deleted by the *garbage collector* (GC). This frees up memory by deleting objects that are no longer used and requesting more memory from the OS.

R uses a tracing GC - is traces every object that's reachable from the global environment and all object reachable from those objects.

GC runs automatically whenever R needs more memory.

Use `gcinfo(TRUE)` and a message is printed whenever GC is run.

Can force gc by calling `gc()`:

```{r}
gc()
```
Not really useful.

`lobstr::mem_used()` is a wrapper around gc() and informs us of the total number of bytes used.

```{r}
mem_used()
```

## Vectors

Will discuss how the different vector types fit together as a whole. 

Vectors are of two types, atomic vectors and lists. The difference relates to their elements' types. Atomic vectors have elements of the same type. List elements can be of different types.

Although not a vector, `NULL` is like a vector and often has the role of a generic zero length vector.

Every vector can also have `attributes` - named list of arbitrary metadata. Two attributes are important, the dimension attribute turns vectors into matrices and arrays and the class attribute powers the S3 object system.

Will discuss important S3 vectors: factors, date and times, data frames and tibbles. R considers data frames and matrices to be vectors (although these 2D structures may not be what comes to mind when thinking about vectors).

### Atomic vectors

*There are four primary types of atomic vectors:* **logical, integer, double and character (which contains strings).** 

Together, integer and double vectors are known as numeric vectors.

There are two rare types, complex and raw - not discussed further here.

#### Scalars

Each of the four primary types has a special syntax to create an individual value - a **scalar**.

 - Logicals can be written in full `TRUE` or `FALSE`, or abbreviated `T` or `F`.
 - Doubles can be specified in decimal (`0.1234`), scientific (`1.23e4`), or hexadecimal (`0xcafe`) form. There are three special values unique to doubles: `Inf`,
   `-Inf` and `NaN` (not a number).
 - Integers are written similarly to doubles, but are followed by an `L`, and cannot contain fractional values.
 - Strings are encompassed by `"` or `'` e.g. `"hi"` or `'bye'`. Special characters are escaped with `\`. `?Quotes` provides further details.
 
#### Making longer vectors with c()

To make long vectors from shorter ones, use c(), which is short for combine:

```{r}
lgl_var <- c(TRUE, FALSE)
int_var <- c(1L, 6L, 10L)
dbl_var <- c(1, 2.5, 4.5)
chr_var <- c("these are", "some strings")
```

When the inputs are atomic vectors, c() always creates another atomic vector, in other words, it flattens:

```{r}
c(c(1,2), c(3, 4))
```

Use `typeof()` and `length()` to determine the type and length of the vector, respectively.

```{r}
typeof(lgl_var)
#> [1] "logical"
typeof(int_var)
#> [1] "integer"
typeof(dbl_var)
#> [1] "double"
typeof(chr_var)
#> [1] "character"
```

#### Missing values

R represents missing or unknown values with `NA` (not applicable). Most computations that contain a missing value will return a missing value:

```{r}
NA > 5
10 * NA

```

A common mistake when identifying missing values in a vector:

```{r}
x <- c(NA, 5, NA, 10)
x == NA
#> [1] NA NA NA NA
```
You should use is.na() to test for the presence of a missing value:

```{r}
is.na(x)
```

There are actually four missing values, one for each of the atomic types: `NA` (logical), `NA_integer_` (integer), `NA_real_` (double), and `NA_character_` (character). 

#### Testing and coercion

You can test if a vector is of a given type with an `is.*()` function. However, use these with care. The following can be used and work as expected:

 - `is.logical()`
 - `is.integer()`
 - `is.double()`
 - `is.character()`

Don't use `is.vector()`, `is.atomic()`, and `is.numeric()`, they don't work as expected.

For atomic vectors, type is a property of the entire vector: all elements must be the same type. When you attempt to combine different types they will be coerced in a fixed order: character → double → integer → logical. For example, combining a character and an integer yields a character:

```{r}
str(c("a", 1))
#>  chr [1:2] "a" "1"
```

Above, the 1 is coerced into a character.

This type of coercion occurs automatically. Many mathematical functions (`+`, `log` etc) coerce to numeric. This is helpful with logical vectors, as `TRUE` becomes `1` and `FALSE` becomes `0`.

```{r}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)
#> [1] 0 0 1

# Total number of TRUEs
sum(x)
#> [1] 1

# Proportion that are TRUE
mean(x)
#> [1] 0.333
```
You can intentionally coerce by using as `as.*()` function, such as `as.logical()`, `as.integer()`, `as.double()`, or `as.character()`. The failed coercion of strings generates a warning and a missing value.

```{r}
as.integer(c("1", "1.5", "a"))
#> Warning: NAs introduced by coercion
#> [1]  1  1 NA
```

#### Exercises

 - vector coercion rules:
 
```{r}
c(1, FALSE)
# 1 0 coerced into double
c("a", 1)
# "a" "1" coerced into character
c(TRUE, 1L)
# 1 1 coerced into double
```

```{r}
1 == "1"
# TRUE as "1" is coerced into a numeric
-1 < FALSE
# TRUE as FALSE is coerced into a numeric and becomes 0
"one" < 2
# FALSE as appears as though 2 is coerced into character

```

### Attributes

Notice that the aforementioned set of atomic vectors doesn't include important data structures like matrices, arrays, factors, or date-times. These types are built on top of atomic vectors by adding attributes. Here will discuss how the dim attribute makes matrices and arrays, and following this, will discuss how the class attribute creates S3 vectors, including factors, dates and date-times.

#### Getting and setting

You can think of attributes as name-value pairs - they attach metadata to an object. Individual attributes can be retrieved and modified  with `attr()`, or retrieved en masse with `attributes()`, and set en masse with `structure()`.

```{r}
a <- 1:3
# set attribute
attr(a, "x") <- "abcdef"
# get attribute
attr(a, "x")

attr(a, "y") <- 4:6
# get all attributes
str(attributes(a))

# Or equivalently

a <- structure(
  1:3,
  x = "abcdef",
  y = 4:6
)

str(attributes(a))

```

Such attributes should generally be thought of as non-permanent i.e most attributes are lost by most operations:

```{r}
attributes(a[1])
# NULL

attributes(sum(a))
# NULL
```

Generally only two attributes preserved:

 - **names**, a character vector giving each element a name
 - **dim**, short for dimension, an integer vector, used to turn vectors into matrices or arrays.
 
To preserve other attributes, we will need to create our own S3 class, discussed later.

#### Names

